---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

echo = TRUE

## Loading and preprocessing the data
```{r}
### Load the data
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url,"data.zip") ##download the zip file
data <- read.csv(unz("data.zip", filename="activity.csv")) ##unzip and read the csv

### Preprocess data
data$date = as.Date(data$date, format = "%Y-%m-%d") ##set to date class
data$interval = as.factor(data$interval)
```


## 1. What is mean total number of steps taken per day?
First the steps by date are loadaded in the *stepsbydate* variable.
Then the data is plotted to a histogram.

```{r}
stepsbydate <- aggregate(steps ~ date, data = data, FUN = sum, 
                        na.action = na.omit) ## sum data by date
head(stepsbydate) ##shows the first data items for stepsbydate
```
1.1 The first of the calculated values of steps by date are shown above.

```{r}
hist(stepsbydate$steps, main = "Histogram of steps by date",
     xlab = "Amount of steps by date") ## Plot a histogram of resulting data
### CAlculate mean and median
mean = mean(stepsbydate$steps)
median = median(stepsbydate$steps)
```
1.2 The histogram for total number of steps per day is shown above  

1.3 The mean is `r mean` steps per day, and the median is `r median` steps by day.

## 2. What is the average daily activity pattern?
The data is aggregated by day to plot average daily values. Note values are divided by number of days.

```{r}
stepsbytime <- aggregate(steps ~ interval, data = data, FUN = sum)

### Sum by date
stepsbytime$meansteps = stepsbytime$steps / nrow(stepsbydate)
plot(stepsbytime$interval, stepsbytime$meansteps, xlab = "Interval",
    ylab = "Steps averaged by day", main = "Daily Activity Pattern", type = "l")
lines(stepsbytime$interval, stepsbytime$meansteps)
maxint = which.max(stepsbytime$meansteps)
value = max(stepsbytime$meansteps)
```
2.1 The Interval plot is shown above  

2.2 The interval with the maximum number of steps is interval `r maxint`, which has `r value` steps on average.


## 3. Imputing missing values
First we report the number of missing values with the code: 
```{r}
missing = sum(is.na(data))
```
1.1 There are `r missing` missing values.

1.2 Then we segregate the NA values from the data set, assign average interval values to the missing steps values and merge the data with the original data.
```{r}
library(plyr) ##for using mapvalues function
## mapvalues function is used to easily write the average steps values to each missing entry given the interval number for the entry.

NAs <- data[is.na(data),] ## Segregate NA values
NAs$steps <- mapvalues(NAs$interval, from = levels(data$interval), to = stepsbytime$meansteps) ## return average step values for given interval in "steps" column
NAs$steps = as.numeric(as.character(NAs$steps)) ##save as numeric values
dat2 = data ##create "dat2" variable to contain given and replaced missing vals.
dat2[is.na(data),1] = NAs$steps ##fill missing values
head(dat2)
```
1.3 As we can observe, missing values have been replaced with average interval values. We can now re-evaluate the data like previously done.

```{r}
stepsbydate <- aggregate(steps ~ date, data = dat2, FUN = sum, na.action = na.omit) ## sum by date
hist(stepsbydate$steps, main = "Histogram of steps by date", 
     xlab = "Amount of steps by date") ##Plot a histogram of the resulting data
## 
mean = mean(stepsbydate$steps)
median = median(stepsbydate$steps)
```

1.4 The mean is `r mean` and the median is `r median`.


## 4. Are there differences in activity patterns between weekdays and weekends?
1.1 First, the weekend/weekday factor is created by using weekdays() function.

```{r}

days <- weekdays(dat2$date)##Give name of day
wkend <- c("Saturday", "Sunday")##weekend days
wtest <- days %in% wkend##logical value of weekend or not
dat2$WE <- factor(wtest, labels = c("Weekday","Weekend")) ##Weekend factor var.

```

1.2 Then we plot the results.

```{r}
## Work with steps by date data to count number of weekdays and weekend days.
idays <- weekdays(stepsbydate$date) ##List of individual day names
n <- length(idays) ## Total number of days
w <- sum(idays %in% wkend) ## Number of weekend days
d <- n-w ## Number of weekdays

## Separate filled in data frame into weekdays (d) and weekend days(w).
datw <- dat2[wtest,] ## Weekend dataset
datd <- dat2[!wtest,] ## Weekday dataset

## Aggregate weekend data for plotting
wstepsbytime <- aggregate(steps ~ interval, data = datw, FUN = sum)
wstepsbytime$meansteps <- wstepsbytime$steps / w

## Aggregate weekday data for plotting
dstepsbytime <- aggregate(steps ~ interval, data = datd, FUN = sum) 
dstepsbytime$meansteps <- dstepsbytime$steps / d

## Generate plot
par(mfrow = c(2,1))
plot(wstepsbytime$interval, wstepsbytime$meansteps, type = "l", 
     xlab = "Interval", ylab = "Avg. steps", 
     main = "Weekend")
plot(dstepsbytime$interval, dstepsbytime$meansteps, type = "l", 
     xlab = "Interval", ylab = "Avg. steps", 
     main = "Weekday")
dev.off()
```